### 2022-10-12 draft/draft.tex
still writing

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{year,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{apalike}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\setlength{\parindent}{20.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
\newcommand{\SBM}{\mathrm{SBM}}
\newcommand{\DCBM}{\mathrm{DCBM}}
\newcommand{\argmin}{\mathrm{argmin}}

@@

\begin{example}[Graph clustering]
$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
\begin{equation}
\label{eq:kmeans}
W(x_1, ..., x_n) = \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
\end{equation}
in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments we wish to optimize for, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration labels are updated according to which centroid they are closest to and centroids are updated via the sample mean of each cluster.

If instead of vectors in $\mathbb{R}^p$ we have a graph, it is not obvious how to translate $K$-means clustering to these data. 
In particular, there is no inherent notion of centroid or sample mean for graph data. 
However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
One way to adapt $K$-means to graph data is to use an equivalent objective function as equation~\eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:  
$$
\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
$$
Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
If they are vectors in Euclidean space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
or if they are vertices on a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
An alternative algorithm is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.

Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
One way to combine the graph distance approach and the graph embedding approach is to apply Lloyd's algorithm on an embedding that approximates graph distances \citep{vonLuxburg2007}. 
\end{example}

We now provide a more formal description of graph data:
Suppose we observe a network of \(n\) objects and pairwise relationships between them.
This network is represented by a graph \(G = (V, E)\) with vertex set \(V = \{v_1, ..., v_n\}\), representing the \(n\) objects, and edge set \(E\), representing the up to \(n (n - 1) / 2\) pairwise relationships (assuming that there are no self-loops).
The numeric representation of these data is in the form of \emph{affinity matrix} \(A \in \mathbb{R}^{n \times n}\) in which each \(A_{ij}\) represents object \(i\)'s relationship to object \(j\).
We assume that the entries of \(A\) represent affinities or similarities, i.e., the higher the value of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\).
If \(A_{ij} = 0\), then \(i\) has no direct relationship to \(j\).
\(A\) is symmetric if it represents an undirected graph in which the relationship from \(i\) to \(j\) is the same as the relationship from \(j\) to \(i\).
\(A\) is binary, i.e., \(A \in \{0, 1\}^{n \times n}\), if it represents an unweighted graph in which edges either exist or don't exist.
If \(A\) is binary, we call it an \emph{adjacency matrix}.
In this work, we focus primarily on undirected and unweighted graphs without self loops.

@@

Within the scope of this work, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs.
The adjacency matrix that describes these graphs is binary, symmetric, and hollow.
In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., \(A_{ij} \indep \Bernoulli(P_{ij})\) for some \(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\) since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
Then similar to how the edges are compiled into an adjacency matrix \(A\), the edge probabilities can be compiled into an edge probability matrix \(P \in [0, 1]^{n \times n}\).
This type of graph model is defined as a \emph{Bernoulli graph} (also called an inhomogeneous \(\ER\) graph).
If \(A\) is the adjacency matrix of a Bernoulli graph with edge probability matrix \(P\), we denote \(A \sim \BG(P)\).
If the vertices and edge probabilities are sampled as a sequence of \(n\) i.i.d. random variables from probability distribution \(F\), then we denote \(A \sim \BG(F, n)\).

Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
Additional structure must be introduced.
One such structured Bernoulli graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:

@@

\(\ER\) graphs lie on the opposite end of the spectrum in that they restrict the model to a single parameter, which does not result in a very interesting model or reflect many networks observed in real data.
Much work has been done in developing various Bernoulli graph models that strike a balance between structure and flexibility.
Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called \emph{block models}, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called \emph{latent space models}.

\textbf{{[}Maybe insert something here about the exponential family of random graph models and how, like mixture distributions, block models do not fit within this family?{]}}

@@

Network analysis is often concerned with community detection.
This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices \(i\) and \(j\) depend on the communities to which vertices \(v_i\) and \(v_j\) belong.
More formally, one assigns each vertex \(v_i\) a community label \(z_i\) and assumes a Bernoulli graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
Within the context of this work, we will call such models \emph{block models}.
This type of model framework allows us to frame the goal of community detection as a statistical inference problem:
identify the true community labels, up to permutation of the labels.

@@

The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities \((k, \ell)\) has a constant edge probability \(\theta_{k \ell}\).
We now give the formal definition of the SBM within the context of Bernoulli graphs:

\begin{definition}[Stochastic block model]
\label{def:sbm}
Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$). 
Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
If the edge probability matrix $P$ for this graph is such that 
$P_{ij} = \theta_{z_i, z_j}$ and $A \sim \BG(P)$, then $G$ is a stochastic block model. 
\end{definition}

@@

An extension to these include Bayesian models, e.g., each \(\theta_{k \ell}\) is sampled from a distribution with support \([0, 1]\).

@@

In most conceptions of block models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}.
For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the network appears to confirm this.
However, SBMs are not necessarily restricted to this type of community structure.
We could just as easily construct an example in which \(\theta_{11}\) and \(\theta_{22}\) are less than \(\theta_{12}\).
Within the context of block models, we say that a graph is \emph{assortative} if \(P\) is positive semidefinite and \emph{disassortative} otherwise, which roughly correspond to communities with higher within-community edge probabilities and lower between-community edge probabilities, respectively.
For example, the SBM in example \ref{ex:assort-sbm} is assortative and has edge probability matrix \(P\) that is rank 2 with nonzero eigenvalues \(n (\frac{3}{8} \pm \frac{\sqrt{2}}{8}) > 0\).
In the following example, we describe a network which can be modeled as a disassortative SBM.

@@

The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard.
Because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the expectation maximization (EM) algorithm \citep{10.2307/2984875}.
However, due to the \(z_{ik} z_{j \ell}\) interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}.
While the labels are drawn independently, they are not necessarily conditionally independent.
Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form.
As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.

@@

So far, we have described \emph{dense} Bernoulli graphs.
For such models, as the sample size \(n\) increases, the expected degree of each node grows linearly.
For example, consider the SBM with the third sampling setup described in section \ref{sec:sbm}.
Then the expected degree of vertex \(v_i\) is
\[
\begin{split}
E[d_i] & = \sum_j P_{ij} \\
& = \sum_{j \neq i} \theta_{z_i, z_j} \\ 
& = \Big( \sum_k n_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
& = n \Big( \sum_k \alpha_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
& = O(n).
\end{split}
\]

@@

The SBM assumes homogeneity within each community.
In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label.
The Degree Corrected Block Model (DCBM) and Popularity Adjusted Block Model (PABM) are generalizations of the SBM that address these limitations.
In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label.
We also show that the DCBM is a special case of the PABM and that the SBM, DCBM, and PABM are nested models.

\begin{definition}[Degree corrected block model]
\label{def:dcbm}
Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM. 
Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
In addition, each vertex $v_i$ has a degree correction parameter $\omega_i \in [0, 1]$. 
If the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model. 
We use the notation $A \sim \DCBM(z, \{\theta_{k \ell}\}_K, \omega; \rho_n)$ to denote a random adjacency matrix drawn from the DCBM with labels $z$, base edge probabilities $\{\theta_{k \ell}\}_K$, and degree correction factors $\omega$. 
Under the sampling scheme in which the labels are drawn from a multivariate distribution with class probabilities $\alpha$, we use the notation $A \sim \DCBM(\alpha, \{\theta_{k \ell}\}_K, \omega; \rho_n)$.
\end{definition}

Showing explicitly that the SBM is a special case of the DCBM only requires us to set \(\omega\) as a constant vector, e.g., \(\omega = \begin{bmatrix} 1 & \cdots & 1 \end{bmatrix}^\top\).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{year,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{apalike}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\setlength{\parindent}{20.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
\newcommand{\SBM}{\mathrm{SBM}}
\newcommand{\DCBM}{\mathrm{DCBM}}

@@

\begin{example}[Graph clustering]
$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
\begin{equation}
\label{eq:kmeans}
\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ W(z_1, ..., z_n; x_1, ..., x_n) = 
\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
\end{equation}
in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments which we wish to optimize, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.

If we have a graph instead of vectors in $\mathbb{R}^p$, it is not obvious how to translate $K$-means clustering to these data. 
In particular, there is no inherent notion of centroid or sample mean for graph data. 
However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
One way to adapt $K$-means to graph data is to use an equivalent objective function as equation \eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:  
$$
\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \tilde{W}(z_1, ..., z_n; x_1, ..., x_n) =
\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
$$
Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
If they are vectors in Euclidean space, then we can use $d(x_i, x_j) = \|x_i - x_j\|_2$, 
or if they are vertices on a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
An alternative algorithm is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.

Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
In this context, a natural choice of embedding is one in which pairwise Euclidean distances between embedding vectors are (approximately) equal to pairwise graph distances between corresponding vertices. 
For example, the combinatorial Laplacian eigenmap is an embedding in which Euclidean distances between pairs of vectors in the embedding are equal to expected commute times between corresponding pairs of vertices in the graph \citep{vonLuxburg2007}. 
\end{example}

We now provide a more formal description of graph data:
Suppose we observe a network of \(n\) objects and pairwise relationships between them.
This network is represented by a graph \(G = (V, E)\) with vertex set \(V = \{v_1, ..., v_n\}\), representing the \(n\) objects, and edge set \(E\), representing the up to \(n (n - 1) / 2\) pairwise relationships (assuming that there are no self-loops).
The numeric representation of these data is in the form of \emph{affinity matrix} \(A \in \mathbb{R}_+^{n \times n}\) in which each \(A_{ij}\) represents object \(i\)'s relationship to object \(j\).
We assume that the entries of \(A\) represent affinities or similarities, i.e., the higher the value of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\).
If \(A_{ij} = 0\), then \(i\) has no direct relationship to \(j\).
\(A\) is symmetric if it represents an undirected graph in which the relationship from \(i\) to \(j\) is the same as the relationship from \(j\) to \(i\).
\(A\) is binary, i.e., \(A \in \{0, 1\}^{n \times n}\), if it represents an unweighted graph in which edges either exist or don't exist.
If \(A\) is binary, we call it an \emph{adjacency matrix}.
In this work, we focus primarily on undirected and unweighted graphs without self loops.

@@

Within the scope of this work, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs.
The adjacency matrix that describes these graphs is binary, symmetric, and hollow.
In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., \(A_{ij} \indep \Bernoulli(P_{ij})\) for some \(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\) since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
Then similar to how the edges are compiled into an adjacency matrix \(A\), the edge probabilities can be compiled into an edge probability matrix \(P \in [0, 1]^{n \times n}\).
This type of graph model is defined as a \emph{Bernoulli random graph} (also called an inhomogeneous \(\ER\) graph).
If \(A\) is the adjacency matrix of a Bernoulli random graph with edge probability matrix \(P\), we denote \(A \sim \BG(P)\).
If the vertices and edge probabilities are sampled as a sequence of \(n\) i.i.d. random variables from probability distribution \(F\), then we denote \(A \sim \BG(F, n)\).

Statistical inference is not possible on a general Bernoulli random graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
Additional structure must be introduced.
One such structured Bernoulli random graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:

@@

\(\ER\) graphs lie on the opposite end of the spectrum in that they restrict the model to a single parameter, which does not result in a very interesting model or reflect many networks observed in real data.
Much work has been done in developing various Bernoulli random graph models that strike a balance between structure and flexibility.
Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called \emph{block models}, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called \emph{latent space models}.

@@

Network analysis is often concerned with community detection.
This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices \(v_i\) and \(v_j\) depend on the community assignments \(z_i\) and \(z_j\).
More formally, one assigns each vertex \(v_i\) a community label \(z_i \in \{1, ..., K\}\) and assumes a Bernoulli random graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
Within the context of this work, we will call such models \emph{block models}.
This type of model framework allows us to frame the goal of community detection as a statistical inference problem:
come up with estimators for the true community labels, up to permutation.

@@

The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities \((k, \ell)\) has a constant edge probability \(\theta_{k \ell}\).
We now give the formal definition of the SBM within the context of Bernoulli random graphs:

\begin{definition}[Stochastic block model]
\label{def:sbm}
Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$). 
Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
If the edge probability matrix $P$ for this graph is such that 
$P_{ij} = \theta_{z_i, z_j}$ and $A \sim \BG(P)$, then $G$ is a stochastic block model. 
\end{definition}

@@

An extension to these include Bayesian models in which there are prior distributions for the parameters, e.g., each \(\theta_{k \ell}\) has a prior distribution with support \([0, 1]\), such as \(\theta_{k \ell} \sim \Betadist(a_{k \ell}, b_{k \ell})\), and the mixture parameters \(\alpha\) has a prior distribution on the \(K-1\) simplex, such as \(\alpha \sim \mathrm{Dirichlet}(\beta_1, ..., \beta_K)\).

@@

In most conceptions of block models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}.
This reflects the general intuition that pairs within the same community are more likely to be connected than pairs between two different communities.
For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the network appears to confirm this.
However, SBMs are not necessarily restricted to this type of community structure.
We could just as easily construct an example in which \(\theta_{11}\) and \(\theta_{22}\) are less than \(\theta_{12}\).
Within the context of block models, we say that a graph is \emph{assortative} if \(P\) is positive semidefinite and \emph{disassortative} otherwise, which roughly correspond to communities with higher within-community edge probabilities and lower between-community edge probabilities, respectively.
For example, the SBM in example \ref{ex:assort-sbm} is assortative and has edge probability matrix \(P\) that is rank 2 with nonzero eigenvalues \(n (\frac{3}{8} \pm \frac{\sqrt{2}}{8}) > 0\).
In the following example, we describe a network which can be modeled as a disassortative SBM.

@@

The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard.
But because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the widely studied expectation maximization (EM) algorithm \citep{10.2307/2984875}.
However, due to the \(z_{ik} z_{j \ell}\) interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}.
While the labels are drawn independently, they are not necessarily conditionally independent.
Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form.
As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.

@@

So far, we have described \emph{dense} Bernoulli random graphs.
For such models, as the sample size \(n\) increases, the expected degree of each node grows linearly.
For example, consider the SBM with the third sampling setup described in section \ref{sec:sbm}.
Then the expected degree of vertex \(v_i\) is
\[
\begin{split}
E[d_i] & = \sum_j P_{ij} \\
& = \sum_{j \neq i} \theta_{z_i, z_j} \\ 
& = \Big( \sum_k n_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
& = n \Big( \sum_k \alpha_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
& = O(n).
\end{split}
\]

@@

The SBM assumes homogeneity within each community.
In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label.
The degree corrected block model (DCBM) and popularity adjusted block model (PABM) are generalizations of the SBM that address these limitations.
In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label.
We also show that the DCBM is a special case of the PABM and that the SBM, DCBM, and PABM are nested models.

\begin{definition}[Degree corrected block model]
\label{def:dcbm}
Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM. 
Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
In addition, each vertex $v_i$ has a degree correction parameter $\omega_i \in [0, 1]$. 
If the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model. 
We use the notation $A \sim \DCBM(z, \{\theta_{k \ell}\}_K, \omega; \rho_n)$ to denote a random adjacency matrix drawn from the DCBM with labels $z$, base edge probabilities $\{\theta_{k \ell}\}_K$, and degree correction factors $\omega$. 
Under the sampling scheme in which the labels are drawn from a multivariate distribution with class probabilities $\alpha$, we use the notation $A \sim \DCBM(\alpha, \{\theta_{k \ell}\}_K, \omega; \rho_n)$.
\end{definition}
